Wait-free Queues with Polylogarithmic Step Complexity

### 2022-10-04 thesis/thesis.tex
finalize text

\usepackage[letterpaper, margin=1in]{geometry}
\usepackage{tikz-qtree}
\usepackage{algorithm}
\usepackage{algpseudocode}
\makeatletter
\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
\makeatother
\usepackage{graphicx}
\usepackage{subcaption}
%\usepackage{showkeys}
\usepackage{hyperref}
\usepackage{amsmath}

@@

\Function{boolean}{Refresh()}{}
\State \tt{h= \head} \label{readHead}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
\If{\nf{dir.blocks[h\sub{dir}]!=null}} \label{ifHeadnotNull}
\State{\tt{dir.\Call{Advance}{h\sub{dir}, h}}} \label{helpAdvance}
\EndIf
\EndFor \label{endHelpChild1}
\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} 
\EndIf
\State{\tt{result= blocks[h].CAS(null, new)}} \label{cas}

\State \tt{h\sub{p}= parent.\head} \label{readParentHead}
\State{\tt{this.\Call{Advance}{h, h\sub{p}}}} \label{advance}
\State \Return{ \tt{result}}

@@

\Function{void}{Advance}{\sl{int} h, \sl{int} hp}

\State \tt{blocks[h].super.CAS(null, hp)} \label{setSuper1}
\State \tt{head.CAS(h, h+1)} \label{incrementHead}
\EndFunction{Advance}

@@

The next two observations are correct for both enqueues and deqeueues.

\begin{observation}
  $n\nf{.blocks[}b\nf{]}\nf{.num\sub{enq}}$ is the size of $E(n,b)$.
\end{observation}

\begin{observation}\label{lem::sumDeq}
$\sum_{i=1}^{b} {E(n,i)}=n\nf{.blocks[}i\nf{].sum\sub{enq}}$.
\end{observation}

The next observation is correct for both directions.

\begin{observation}\label{lem::numEnqLeft}
Let $B$ be $n\nf{.blocks[}b\nf{]}$. $B\nf{.num\sub{enq-left}}$ is the number of \nf{Enqueue} operations in the direct subblocks of $B$ in $n$\nf{.left}.
\end{observation}
\begin{proof}
Let $B^\prime$ be $n\nf{.blocks[}b-1\nf{]}$. We prove the claim by induction on height of node $n$. 
\begin{align*}
    B\nf{.num\sub{enq-left}}=&B.\nf{sum\sub{enq-left}}-B^\prime.\nf{sum\sub{enq-left}}& \text{Definition of }\nf{num\sub{x}}\\
    =&B^\prime.\nf{sum\sub{enq-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}\\&-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}-B^\prime.\nf{sum\sub{enq-left}} & \nf{CreateBlock} \text{ code}\\
    =&\sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{enq}}
\end{align*}
\end{proof}
\pagebreak

@@

\nf{DSearch($e$, $end$)} returns a pair \nf{<$b$, $i$>} so that the $i$th \nf{Enqueue} in the $b$th block of the root is the $e$th \nf{Enqueue} in the entire sequence stored in the root. 
\begin{lemma}[\nf{DSearch} Correctness]\label{dsearch}
If $\nf{root.blocks[$end$]}\neq \nf{null}$ and $1\leq e \leq \nf{root.blocks[$end$].sum\sub{enq}}$, \nf{DSearch($e$, $end$)} returns \nf{<$b$, $i$>} such that $E_i(root,b)= E_e(root)$.
\end{lemma}
\begin{proof}
\nf{DSearch} performs a doubling search from \nf{root.blocks[end]} to \nf{root.blocks[0]} to find $E_e(root)$. From Lines \ref{setSumEnqLeft}, \ref{setSumEnqRight} we know the \nf{sum\sub{enq-left}} and \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order. Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} is also non-decreasing. Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b$].sum\sub{enq}} \geq e$ and $\nf{root.blocks[$b-1$].sum\sub{enq}} < e$ by Lemma \ref{lem::sumDeq}. Block \nf{root.blocks[$b$]} contains $E_i(root,b)$. The doubling search on Lines \ref{dsearchStart}--\ref{dsearchEnd} doubles its search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[end].sum\sub{enq}}$. In Line \ref{dsearchBinarySearch}, the binary search finds the $b$ required in the range mentioned. Finally $i$, is computed from the definition of \nf{sum\sub{enq}} and lemma \ref{lem::sumDeq}.
\end{proof}

@@

For the induction step we prove if \nf{$n$.child.GetEnqueue($b^\prime$, $i$)} returns $E_i(n.child, b^\prime)$ then \nf{$n$.GetEnqueue($b$,~$i$)} returns $E_i(n,b)$.  From Definition \ref{ordering} of $E(n,b)$, operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}). By Observation \ref{lem::numEnqLeft}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue()} operations from the blocks's subblocks in the left child of \nf{n}. So the $i$th \tt{enqueue} operation in $n\nf{.blocks[}b\nf{]}$ is propagated from the right child if and only if $i$ is greater than $n\nf{.blocks[}b\nf{]}$\nf{.num\sub{enq-left}}. Line \ref{leftOrRight} decides whether the \nf{i}th enqueue in  $b$th block  of internal node $n$ is in the  left child or right child subblocks of \nf{$n$.blocks[b]}. By Definitions \ref{def::ops} and \ref{def::subblock} to find an operation in the subblocks of \nf{$n$.blocks[$i$]} we need to search in the range
\begin{align*}
&\texttt{$n$.left.blocks[$n$.blocks[$i$-1].end\textsubscript{left}+1..$n$.blocks[$i$].end\textsubscript{left}]}  \textrm{ or } \\
&\texttt{$n$.right.blocks[$n$.blocks[$i$-1].end\textsubscript{right}+1..$n$.blocks[$i$].end\textsubscript{right}]}.
 \end{align*}
First we consider the case where the \nf{Enqueue} we are looking for is in the left child. There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ \tt{enqueue}s in the blocks before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.\tt{left})$ which is $E_{i^\prime}(n.\tt{left},b^\prime)$ for some $b^\prime$ and $i^\prime$. We can compute $b^\prime$ and then search for the $i+eb$th enqueue in \tt{n.left}, where $i^\prime$ is $i+eb-n$\tt{.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}. The parameters in Line \ref{leftChildGet} are for searching $E_{i+eb}(n.left)$ in \nf{$n$.left.blocks} in the range of left subblocks of $n$\nf{.blocks[}$b$\nf{]}, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.

@@

\begin{lemma}
Let $R_n$ be the \nf{Refresh} that puts $B$ in \nf{$n$.blocks[$b$]} at Line \ref{cas}. Then, the block created by one of the next two successful \nf{$n$.parent.Refresh}es according to Definition \ref{orderRefresh} contains $B$ and \nf{$B$.super} is set before Line \ref{invokeCreateBlock} of the the second successful \nf{$n$.parent.Refresh}.
\end{lemma}

\begin{proof}
Let the first successful \nf{$n$.parent.Refresh} after $R_n$ be $R_{p1}$ and the second next successful \nf{$n$.parent.Refresh} be $R_{p2}$. To derive a contradiction assume $B$ was not propagated to $n$\nf{.parent} neither by $R_{p1}$ nor by $R_{p2}$.

@@

$R_n$ puts $B$ into $n$\nf{.blocks[}$b$\nf{]} so $n.$\nf{head} is $b$ at $_{\ref{cas}}^{R_n}t$. Since $R_{p2}$'s \nf{CAS} into $n$\nf{.blocks} is successful there should be a \nf{Refresh} instance $R_p^\prime$ on $n$\nf{.parent} that increments $n$\nf{.parent} after $R_{p1}$'s \ref{cas}. $R_{p2}$'s \ref{readHead} is after $R_{p2}$'s \ref{incrementHead}. $R_p^\prime$'s \ref{incrementHead} is after $R_{p1}$'s \nf{CAS}. We assumed $t^{R_n}_{\ref{cas}}<t^{R_{p1}}_{\ref{cas}}<t^{R_{p2}}_{\ref{cas}}$. Since \ref{readChildHead} is after \ref{readHead} inc the code, $R_{p2}$ read a value $\geq b$ for $n.\nf{head}$.

@@

Furthermore, if $B$\nf{.super} is not setset earlier it is set by $R_{p2}$ in Line \ref{setSuper1} invoked from \ref{helpAdvance}.

@@

\begin{figure}[hbt]  
  \center\includegraphics[width=5.5in,height=0.65in]{pics/timeLineSuper1.png}
  \caption{Time relations between \nf{R\sub{n},R\sub{p},R',R"}}
  \label{fig::timeLine1}
\end{figure}

@@

\begin{figure}[hbt]  
  \center\includegraphics[width=6in,height=1.25in]{pics/timeLineSuper2.png}
  \caption{$R_{p2}$ reads \nf{$n$.parent.head} after $t_{320}^{R_{p1}}$, which is after $t_{320}^{Rn}$. $R_{p2}$ has to help increment \nf{$n$.head} and set $b$\nf{.super}.}
  \label{fig::timeLine2}
\end{figure}
TODO: update figures

@@

\begin{corollary} \label{superRelation}
If \nf{$n$.blocks[$b$]} is propagated to $n$\nf{.parent}, then \nf{$n$.blocks[$b$].super} is equal to or one less than the index of the superblock of \nf{$n$.blocks[$b$]}.
\end{corollary}
\begin{proof}
\nf{super} field of a block is set after installing it  to the value of the parent node's \nf{head} by \nf{Advance}. From previous Lemma we know that the block is propagated by second next successful \nf{Refresh} on parent, so the value written in $b$.\nf{super} can  the \nf{super} field of a lemma i
\end{proof}
TODO: complete proof

@@

\begin{lemma}[\nf{IndexDequeue} correctness]
 If $1\leq i \leq\nf{$n$.blocks[$b$].num\sub{deq}}$ then \nf{$n$.IndexDequeue($b$,$i$)} returns $<x,y>$ such that $D_i(n,b)=D_y(\nf{root},x)$.
\end{lemma}
\begin{proof}
We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (Line \ref{indexBaseCase}).
For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes $sb$, index of the superblock of the $b$th block in $n$, in Line \ref{computeSuper} by Corollary \ref{superRelation}. After that the position of $D_i(n,b)$ in $D(n.\nf{parent}, sb)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}. By Definition \ref{ordering} \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right. If $D_i(n,b)$ was propagated from the left child, the number of dequeus in the left subblocks of \nf{$n$.parent.blocks[$sb$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{deqBeforeLeft}. Else if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}). Finally \nf{IndexDequeue} is called on \nf{$n$.parent} recursively and it returns the response from induction hypothesis. 
\end{proof}
TODO: add figure

@@

In the next lemma we show that the \nf{size} field in each \nf{root block} is computed correctly.
\begin{lemma}\label{sizeCorrectness}
  \nf{root.blocks[$b$].\size}is the \size of the queue if the operations in \nf{root.blocks[$0\cdots b$]} are applied in the order of~$L$.  
\end{lemma}
\begin{proof}
We prove the claim by induction on $b$. The base case when ${b=0}$ is trivial since the queue is initially empty and $\nf{root.blocks[}0\nf{].size}=0$. We are going to show the correctness when $b=i$ assuming  correctness when $b=i-1$. By Definition \ref{ordering} \tt{enqueue} operations come before \tt{dequeue} operations in a block. By Lemma \ref{lem::sumDeq} if there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].num\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}. Otherwise the size of the queue after the $b$th block in the root is $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- \nf{root.blocks[}b\nf{].num\sub{deq}}$. In both cases, this is same as the assignment on Line \ref{computeLength}.
\end{proof}

The next lemma is useful to compute the number of non-null dequeues.
\begin{lemma} \label{numberOfNND}
If operations in the root are applied with the order of $L$, the number of non-null \tt{dequeue}s in $\nf{root.blocks[}0\cdots b\nf{]}$ is \nf{root.blocks[$b$].sum\sub{enq} $-$ root.blocks[$b$].size}.
\end{lemma}
\begin{proof}
There are \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} enqueue operations in $\nf{root.blocks[}0\cdots b\nf{]}$. The size of the queue after doing $\nf{root.blocks[}0\cdots b\nf{]}$ in order $L$ is 
the number of $\it{enqueues}$ in $\nf{root.blocks[}0\cdots b\nf{]}$ minus the number of $\it{non-null \tt{dequeue}s}$ in $\nf{root.blocks[}0\cdots b\nf{]}$. By the correctness of the \nf{size} field from Lemma \ref{sizeCorrectness} and \nf{sum\sub{enq}} field from Lemma \ref{lem::sumDeq}, the number of $\it{non-null \tt{dequeue}s}$ is \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} $-$ \nf{root.blocks[$b$]}\nf{.size}. 
\end{proof}

@@

A \nf{Dequeue()} creates a block with null value element, appends it to the tree, computes its order among enqueue operations, and returns the response. The first two part is similar to an \nf{Enqueue} operation. To compute the order of a \nf{dqueue} in $D(n)$ there are some constant steps and \nf{IndexDequeue()} is called. \nf{IndexDequeue} does a search with range $p$ in each level (Lemma \ref{superRange}) which takes $O(log^2 p)$ in the tree. In the \nf{FindResponse()} routine \nf{DSearch()} in the root takes $\Theta(\log$(\nf{root.blocks[b].\size+root.blocks[end].\size}) by Lemma \ref{dsearch}, which is $O(\log$ size of the queue when \nf{enqueue} is invoked$)+\log$ size of the queue when \nf{dequeue} is invoked$)$. Each search in \nf{GetEnqueue()} takes $O(\log p)$ since there are $\leq p$ subblocks in a block (Lemma \ref{subBlocksBound}), so \nf{GetEnqueue()} takes $O(\log^2 p)$ steps.

>>>>

\usepackage[letterpaper, margin=1in]{geometry}
\usepackage{tikz-qtree}
\usepackage{algorithm}
\usepackage{algpseudocode}
\makeatletter
\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
\makeatother
\usepackage{graphicx}
\usepackage{subcaption}
%\usepackage{showkeys}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{relsize}
\usepackage{enumitem}

@@

\Function{boolean}{Refresh()}{}
\State \tt{h= \head} \label{readHead}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
\If{\nf{dir.blocks[h\sub{dir}]!=null}} \label{ifHeadnotNull}
\State{\tt{dir.\Call{Advance}{h\sub{dir}}}} \label{helpAdvance}
\EndIf
\EndFor \label{endHelpChild1}
\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} 
\EndIf
\State{\tt{result= blocks[h].CAS(null, new)}} \label{cas}
\State{\tt{this.\Call{Advance}{h}}} \label{advance}
\State \Return{ \tt{result}}

@@

\Function{void}{Advance}{\sl{int} h}
\State \tt{h\sub{p}= parent.\head} \label{readParentHead}
\State \tt{blocks[h].super.CAS(null, hp)} \label{setSuper1}
\State \tt{head.CAS(h, h+1)} \label{incrementHead}
\EndFunction{Advance}

@@

\begin{observation}\label{sumToNum}
$n.\nf{blocks[}i\nf{].sum\sub{x}}-n.\nf{blocks[}j\nf{].sum\sub{x}}=\mathlarger{\sum}_{k=i+1}^{j}  n\nf{.blocks[}k\nf{].num\sub{x}}$ ($i<j$)
\end{observation}

\begin{lemma}\label{lem::numX}
Let $B$, $B^\prime$ be $n\nf{.blocks[}b\nf{]}$, $n\nf{.blocks[}b-1\nf{]}$ respectively and \nf{x} be in $\{\nf{enq},\nf{deq}\}$.
\begin{enumerate}[label=(\arabic*)]
    \item If $n$ is an internal node $B\nf{.num\sub{x-left}}=\Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$
    \item If $n$ is an internal node $B\nf{.num\sub{x-right}}=\Big|E(n.\nf{right},B^\prime.\nf{end\sub{right}}+1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$
    \item $B\nf{.num\sub{x}}=\Big|E(n,b)\Big|$
\end{enumerate}
\end{lemma}
\begin{proof}
 We prove the claim by induction on height of node $n$. Base case (3) for leaves is trivial. Supposing the claim is true for $n$'s children, we prove the correctness of claim for $n$.
\begin{align*}
    B\nf{.num\sub{x-left}}=&B.\nf{sum\sub{x-left}}-B^\prime.\nf{sum\sub{x-left}}& \text{Definition of }\nf{num\sub{x}}\\
    =&B^\prime.\nf{sum\sub{x-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{x}}\\&-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{x}}-B^\prime.\nf{sum\sub{x-left}} & \nf{CreateBlock}\\
    =& n.\nf{left.blocks[}B\nf{.end\sub{left}].sum\sub{x}}- n.\nf{left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{x}}\\
    =&\sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{x}} &\text{Observation} \ref{sumToNum}
\end{align*}
The last line holds because of the induction hypothesis. (2) is similar to (1). Now we prove (3) starting from the Definition of $E(n,b)$.
\begin{align*} 
E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
\end{align*}
By (1) and (2) we have $\Big|E(n,b)\Big|=B\nf{.num\sub{x-left}}+B\nf{.num\sub{x-right}}=B\nf{.num\sub{x}}$
\end{proof}

\begin{corollary}\label{lem::sumX}
Let $B$ be $n\nf{.blocks[}b\nf{]}$ and \nf{x} be in $\{\nf{enq},\nf{deq}\}$.
\begin{enumerate}[label=(\arabic*)]
    \item If $n$ is an internal node $B\nf{.sum\sub{x-left}}=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$
    \item If $n$ is an internal node $B\nf{.sum\sub{x-right}}=\Big|E(n.\nf{right},1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$
    \item $B\nf{.sum\sub{x}}=\Big|E(n,1)\cdot E(n,2)\cdots E(n,b)\Big|$
\end{enumerate}
\end{corollary}

@@

\nf{DSearch($e$, $end$)} returns a pair \nf{<$b$, $i$>} so that the $i$th \nf{Enqueue} in the $b$th block of the root is the $e$th \nf{Enqueue} in the entire sequence stored in the root. 
\begin{lemma}[\nf{DSearch} Correctness]\label{dsearch}
If $\nf{root.blocks[$end$]}\neq \nf{null}$ and $1\leq e \leq \nf{root.blocks[$end$].sum\sub{enq}}$, \nf{DSearch($e$, $end$)} returns \nf{<$b$, $i$>} such that $E_i(root,b)= E_e(root)$.
\end{lemma}
\begin{proof}
\nf{DSearch} performs a doubling search from \nf{root.blocks[end]} to \nf{root.blocks[0]} to find $E_e(root)$. From Lines \ref{setSumEnqLeft}, \ref{setSumEnqRight} we know the \nf{sum\sub{enq-left}} and \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order. Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} is also non-decreasing. Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b$].sum\sub{enq}} \geq e$ and $\nf{root.blocks[$b-1$].sum\sub{enq}} < e$ by Lemma \ref{lem::sumX}. Block \nf{root.blocks[$b$]} contains $E_i(root,b)$. The doubling search on Lines \ref{dsearchStart}--\ref{dsearchEnd} doubles its search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[end].sum\sub{enq}}$. In Line \ref{dsearchBinarySearch}, the binary search finds the $b$ required in the range mentioned. Finally $i$, is computed from the definition of \nf{sum\sub{enq}} and lemma \ref{lem::sumX}.
\end{proof}

@@

For the induction step we prove if \nf{$n$.child.GetEnqueue($b^\prime$, $i$)} returns $E_i(n.child, b^\prime)$ then \nf{$n$.GetEnqueue($b$,~$i$)} returns $E_i(n,b)$.  From Definition \ref{ordering} of $E(n,b)$, operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}). By Observation \ref{lem::numX}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue()} operations from the blocks's subblocks in the left child of \nf{n}. So the $i$th \tt{enqueue} operation in $n\nf{.blocks[}b\nf{]}$ is propagated from the right child if and only if $i$ is greater than $n\nf{.blocks[}b\nf{]}$\nf{.num\sub{enq-left}}. Line \ref{leftOrRight} decides whether the \nf{i}th enqueue in  $b$th block  of internal node $n$ is in the  left child or right child subblocks of \nf{$n$.blocks[b]}. By Definitions \ref{def::ops} and \ref{def::subblock} to find an operation in the subblocks of \nf{$n$.blocks[$i$]} we need to search in the range
\begin{align*}
&\texttt{$n$.left.blocks[$n$.blocks[$i$-1].end\textsubscript{left}+1..$n$.blocks[$i$].end\textsubscript{left}]}  \textrm{ or } \\
&\texttt{$n$.right.blocks[$n$.blocks[$i$-1].end\textsubscript{right}+1..$n$.blocks[$i$].end\textsubscript{right}]}.
 \end{align*}
First we consider the case where the \nf{Enqueue} we are looking for is in the left child. There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ \tt{enqueue}s in the blocks before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.\tt{left})$ which is $E_{i^\prime}(n.\tt{left},b^\prime)$ for some $b^\prime$ and $i^\prime$. We can compute $b^\prime$ and then search for the $i+eb$th enqueue in \tt{n.left}, where $i^\prime$ is $i+eb-n$\tt{.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}. The parameters in Line \ref{leftChildGet} are for searching $E_{i+eb}(n.left)$ in \nf{$n$.left.blocks} in the range of left subblocks of $n$\nf{.blocks[}$b$\nf{]}, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.

@@

\begin{lemma}
Let $R_n$ be the \nf{Refresh} that puts $B$ in \nf{$n$.blocks[$b$]} at Line \ref{cas}. Then, the block created by one of the next two successful \nf{$n$.parent.Refresh}es according to Definition \ref{orderRefresh} contains $B$ and \nf{$B$.super} is set after Line \ref{helpAdvance} of the the second successful \nf{$n$.parent.Refresh}.
\end{lemma}

\begin{proof}
Let $R_{p1}$ be the first successful \nf{$n$.parent.Refresh} after $R_n$ and $R_{p2}$ be the second next successful \nf{$n$.parent.Refresh}. To derive a contradiction assume $B$ was not propagated to $n$\nf{.parent} neither by $R_{p1}$ nor by $R_{p2}$.

@@

$R_n$ puts $B$ into $n$\nf{.blocks[}$b$\nf{]} so ${R_n}$ reads value $b$ for $n.$\nf{head}. Since $R_{p2}$'s \nf{CAS} into $n$\nf{parent.blocks} is successful there should be a \nf{Refresh} instance $R_p^\prime$ on $n$\nf{.parent} that increments $n$\nf{.parent} after $R_{p1}$'s \ref{cas} and before $R_{p2}$'s \ref{readHead}. We assumed $t^{R_n}_{\ref{cas}}<t^{R_{p1}}_{\ref{cas}}<t^{R_{p2}}_{\ref{cas}}$. Finally, since \ref{readChildHead} is after \ref{readHead} and $R_{p2}$'s \ref{readHead} is after $R_p^\prime$'s \ref{incrementHead}, $R_{p2}$ reads a value $\geq b$ for $n.\nf{head}$.

$$_{\ref{cas}}^{R_n}t<_{\ref{cas}}^{R_p1}<_{\ref{incrementHead}}^{R_p^\prime}t<_{\ref{cas}}^{R_p2}t<_{\ref{readChildHead}}^{R_p2}$$
TODO: draw figure

@@

Furthermore, if $B$\nf{.super} is not set earlier it is set by $R_{p2}$ in Line \ref{setSuper1} invoked from \ref{helpAdvance}.

@@

% \begin{figure}[hbt]  
%   \center\includegraphics[width=5.5in,height=0.65in]{pics/timeLineSuper1.png}
%   \caption{Time relations between \nf{R\sub{n},R\sub{p},R',R"}}
%   \label{fig::timeLine1}
% \end{figure}

@@

% \begin{figure}[hbt]  
%   \center\includegraphics[width=6in,height=1.25in]{pics/timeLineSuper2.png}
%   \caption{$R_{p2}$ reads \nf{$n$.parent.head} after $t_{320}^{R_{p1}}$, which is after $t_{320}^{Rn}$. $R_{p2}$ has to help increment \nf{$n$.head} and set $b$\nf{.super}.}
%   \label{fig::timeLine2}
% \end{figure}

@@

\begin{corollary} \label{superRelation}
If \nf{$B=n$.blocks[$b$]} is propagated to $n$\nf{.parent}, then \nf{$B$.super} is equal to or one less than the index of the superblock of $B$.
\end{corollary}
\begin{proof}
After that $B$ is installed $n$\nf{.parent.head} is read and $B.$\nf{super} field is set to the value read from the parent's \nf{head} (see Lines \ref{readParentHead} and \ref{setSuper1} of\nf{Advance}). From previous Lemma we know that $B$ is propagated by second next successful \nf{Refresh}'s \nf{CAS} on $n$\nf{.parent.blocks}. To summarize we have $n\nf{.parent.head}_{_{\ref{cas}}^{R_p2}t}= n\nf{.parent.head}_{_{\ref{cas}}^{R_p1}t}+1$ and $n\nf{.parent.head}_{_{\ref{cas}}^{R_p1}t}\leq n\nf{.parent.head}_{_{\ref{cas}}^{R_n}t}$, so the value read for $n$\nf{.parent.head} is equal to or one less than the index of the superblock of $B$.
\end{proof}

@@

\begin{lemma}[\nf{IndexDequeue} correctness]
 If $1\leq i \leq\nf{$n$.blocks[$b$].num\sub{deq}}$ then \nf{$n$.IndexDequeue($b$,$i$)} returns $<x,y>$ such that $D_i(n,b)=D_y(\nf{root},x)$.
\end{lemma}
\begin{proof}
We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (Line \ref{indexBaseCase}).
For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes $sb$, index of the superblock of the $b$th block in $n$, in Line \ref{computeSuper} by Corollary \ref{superRelation}. After that the position of $D_i(n,b)$ in $D(n.\nf{parent}, sb)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}. By Definition \ref{ordering} \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right. If $D_i(n,b)$ was propagated from the left child, the number of dequeus in the left subblocks of \nf{$n$.parent.blocks[$sb$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{considerPreviousLeft}. Else if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}). Finally \nf{IndexDequeue} is called on \nf{$n$.parent} recursively and it returns the response from induction hypothesis. 
\end{proof}
TODO: add figure

@@

In the next lemma we show that the \nf{size} field in each \nf{root block} is computed correctly.
\begin{lemma}\label{sizeCorrectness}
  \nf{root.blocks[$b$].\size}is the \size of the queue if the operations in \nf{root.blocks[$0\cdots b$]} are applied in the order of~$L$.  
\end{lemma}
\begin{proof}
We prove the claim by induction on $b$. The base case when ${b=0}$ is trivial since the queue is initially empty and $\nf{root.blocks[}0\nf{].size}=0$. We are going to show the correctness when $b=i$ assuming  correctness when $b=i-1$. By Definition \ref{ordering} \tt{enqueue} operations come before \tt{dequeue} operations in a block. By Lemma \ref{lem::numX} if there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].num\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}. Otherwise the size of the queue after the $b$th block in the root is $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- \nf{root.blocks[}b\nf{].num\sub{deq}}$. In both cases, this is same as the assignment on Line \ref{computeLength}.
\end{proof}

The next lemma is useful to compute the number of non-null dequeues.
\begin{lemma} \label{numberOfNND}
If operations in the root are applied with the order of $L$, the number of non-null \tt{dequeue}s in $\nf{root.blocks[}0\cdots b\nf{]}$ is \nf{root.blocks[$b$].sum\sub{enq} $-$ root.blocks[$b$].size}.
\end{lemma}
\begin{proof}
There are \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} enqueue operations in $\nf{root.blocks[}0\cdots b\nf{]}$. The size of the queue after doing $\nf{root.blocks[}0\cdots b\nf{]}$ in order $L$ is 
the number of $\it{enqueues}$ in $\nf{root.blocks[}0\cdots b\nf{]}$ minus the number of $\it{non-null \tt{dequeue}s}$ in $\nf{root.blocks[}0\cdots b\nf{]}$. By the correctness of the \nf{size} field from Lemma \ref{sizeCorrectness} and \nf{sum\sub{enq}} field from Lemma \ref{lem::numX}, the number of $\it{non-null \tt{dequeue}s}$ is \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} $-$ \nf{root.blocks[$b$]}\nf{.size}. 
\end{proof}

@@

A \nf{Dequeue()} creates a block with null value element, appends it to the tree, computes its order among enqueue operations, and returns the response. The first two part is similar to an \nf{Enqueue} operation. To compute the order of a \nf{dqueue} in $D(n)$ there are some constant steps and \nf{IndexDequeue()} is called. \nf{IndexDequeue} does a search with range $p$ in each level which takes $O(log^2 p)$ in the tree. In the \nf{FindResponse()} routine \nf{DSearch()} in the root takes $\Theta(\log$(\nf{root.blocks[b].\size+root.blocks[end].\size}) by Lemma \ref{dsearch}, which is $O(\log$ size of the queue when \nf{enqueue} is invoked$)+\log$ size of the queue when \nf{dequeue} is invoked$)$. Each search in \nf{GetEnqueue()} takes $O(\log p)$ since there are $\leq p$ subblocks in a block (Lemma \ref{subBlocksBound}), so \nf{GetEnqueue()} takes $O(\log^2 p)$ steps.
